---
title: "Quiz 1"
author: "Menachem Sokolik and Idan Keipour"
date: "12 5 2021"
output:
    rmarkdown::github_document:
    theme: journal
    toc: true
    toc_depth: 3
    df_print: paged
---
```{r, cache=TRUE}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

knitr::opts_chunk$set(warning=FALSE)
```

```{r setup, include=FALSE}
library(dummies)
library(dplyr)
library("ggplot2")        
library("GGally")
library(corrplot)

```

```{r}
My_lm <- function(X,Y){
  #' @param X is matrix with n rows and p+1 columns
  #' @param Y is vector with length n 
  # TODO: Complete the code.
  #       Don't use any external libaries or functions for computing linear models.
  X <- as.matrix(X)
  X <- cbind(1,X) # n*(p+1)
  Y <- as.matrix(Y) 
  n <- nrow(X)
  p <- ncol(X) - 1
  dof <-  n-p-1
  # t(X) is the transponse, %*% is matrix multiplication,
  # solve takes the inverse
  A <- solve(t(X) %*% X)
  # this comes from looking at formula for beta hat and
  # multiplying the additional terms needed
  beta_hat <- A %*% t(X) %*% Y
  Y_hat <- X %*% beta_hat
  # just filling in formula
  residuals <- Y - (Y_hat)
  sigma2_hat <- t(residuals) %*% residuals/dof
  # I’ve put as.numeric() at the end to convert it to a single scalar number 
  # instead of a 1 by 1 matrix.
  sigma2_hat <- as.numeric(sigma2_hat)
  # covariance matrix of estimated regression coefficients
  # is just the estimated residual variance times solve(t(X) %*% X)
  # we calculuted earlier and stored as A
  beta_cov <- sigma2_hat * A
  # diag takes the diagonal of the matrix, sqrt makes it go
  # from variance to standard deviation
  beta_SE <- sqrt(diag(beta_cov))
  
  T_stat <- beta_hat/beta_SE
  P_Values <- 2*(1-pt(abs(T_stat),dof))
    
  # residual sum of squares divided by total sum of squares
  R_squared <- 1- sum((residuals)^2)/sum((Y - mean(Y))^2)
  return(list(beta_hat, beta_cov, T_stat, P_Values, sigma2_hat, R_squared, dof, residuals, X, Y, Y_hat))
}
```

```{r}
df <- read.csv("startups.csv")
```
### i
```{r}
str(df)
```
```{r}
summary(df)
```
### ii
```{r}
hist(df$Profit, xlab = "Profit")
```
looks like a normal distribution but with long right tail. 
also can see that with the QQPlot to identify normality.
```{r}
qqnorm(df$Profit)
qqline(df$Profit, lwd = 2)
```
### iii
```{r}
# automatic for the correlation coefficients.
# The diagonal consists of the densities of the three variables and the upper panels consist of the correlation coefficients between the variables.
ggpairs(df[c(-4)])
pairs(~ R.D.Spend + Administration + Marketing.Spend + Profit, data = df, upper.panel = NULL)
```

```{r}
# by hand
cor_df <- cor(df[c(-4)])
corrplot(cor_df, method="number",type = "lower" )
cor_df
```
### iv
```{r}
# matrix of profit
# dummy <- for k variables making k-1 categorical.
df$New_York <- ifelse(df$State == "New York", 1,0)
df$Florida <- ifelse(df$State == "Florida", 1,0)
Y <- df["Profit"] 
X <- df %>% select(-c("Profit","State"))
My_reg <- My_lm(X,Y)
coeff <- My_reg[[1]]
rownames(coeff)[1] = "Intercept"
coeff
```

### v
```{r}
t_stat <- My_reg[3]
t_stat
```
### vi
```{r}
CI <- function(alpha, coeff, beta_cov, dof){
  CI_betas <- data.frame(beta=NA, lower_beta=NA, upper_beta=NA)
  for(i in 1:length(coeff)){
  beta <- rownames(coeff)[i]
  lower_beta <- coeff[i] - sqrt(beta_cov[i])*qt(1-alpha/2,dof)
  upper_beta <- coeff[i] + sqrt(beta_cov[i])*qt(1-alpha/2,dof)
  CI_betas <-rbind(CI_betas, c(beta, lower_beta, upper_beta))}
  return(CI_betas)}
```
### vii
```{r}
alpha= 0.05
beta_cov <- diag(My_reg[[2]])
dof <- My_reg[[7]]
CI(alpha, coeff, beta_cov, dof)
```
We will be rejected H_0 when T_j in is absolute value is greater than qt(dof).
So when truth comes out we reject the null hypothesis otherwise we do not reject.
Also, can not see by p.value test when p.value greater than alpha does not reject the null hypothesis, otherwise rejected.
rejecting null hypothesis of intercept and for R.D spend.
```{r}
abs(t_stat[[1]])>qt(0.975, dof)
P.value <- My_reg[[4]]
P.value
```
## question 4.c
```{r}
new_df <- read.csv("Startups_test.csv")
new_df <- cbind("Intercept"=1,new_df)
new_df$New_York <- ifelse(new_df$State == "New York", 1,0)
new_df$Florida <- ifelse(new_df$State == "Florida", 1,0)
real_profit <- new_df$Profit
new_df<- new_df %>% select(-c("Profit","State"))
new_df <- as.matrix(mutate(new_df,"exp_profit" = as.matrix(new_df)%*%coeff))
new_df
n <- length(real_profit)
RMSE <- sqrt((1/n)*sum((new_df[,7] - real_profit)^2))
RMSE
```